---
title: "How the RePORT-AI PHI Pipeline Works"
subtitle: "A plain-language walkthrough for lab members, with the evidence behind it"
date: "Generated `r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    df_print: kable
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center", fig.width = 9, fig.height = 4.6, dpi = 130)
library(yaml)
library(jsonlite)
library(dplyr)
library(tidyr)
library(ggplot2)

NAVY  <- "#1b2a4a"; TEAL <- "#0f9d8f"; CORAL <- "#e8604c"
GREY  <- "#9aa5b1"; GOLD <- "#e0a72e"

theme_lab <- theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold", color = NAVY, size = 14),
        plot.subtitle = element_text(color = "#5b6472", size = 10.5),
        panel.grid.minor = element_blank(),
        axis.title = element_text(color = "#3a4250"),
        plot.background = element_rect(fill = "white", color = NA))
```

```{css, echo=FALSE}
body { max-width: 1100px; }
.stat-row { display: flex; gap: 14px; flex-wrap: wrap; margin: 18px 0 26px 0; }
.stat-card { flex: 1 1 150px; background: #f4f6f8; border-left: 5px solid #0f9d8f;
  border-radius: 6px; padding: 14px 16px; }
.stat-card .num { font-size: 1.9em; font-weight: 800; color: #1b2a4a; line-height: 1.1; }
.stat-card .lbl { font-size: 0.82em; color: #55606e; margin-top: 4px; }
.callout { border-radius: 6px; padding: 14px 18px; margin: 16px 0; border-left: 6px solid; }
.callout-info    { background: #eaf3fb; border-color: #2f80ed; }
.callout-success { background: #e6f7f3; border-color: #0f9d8f; }
.callout-warn    { background: #fff6e6; border-color: #e0a72e; }
.callout-danger  { background: #fbeceb; border-color: #c0392b; }
.callout b.tag { text-transform: uppercase; font-size: 0.72em; letter-spacing: 0.05em; }
.pipeline-flow { display:flex; align-items:center; flex-wrap:wrap; gap:6px; margin:18px 0; font-size:0.86em;}
.pflow-box { border-radius:8px; padding:10px 12px; min-width:118px; text-align:center; font-weight:600; color:white; }
.pflow-zone-header{ font-weight:800; font-size:0.78em; color:#5b6472; text-transform:uppercase; letter-spacing:.06em; width:100%; margin:10px 0 4px 0;}
.pflow-arrow { font-size:1.3em; color:#9aa5b1; }
.capmatrix { width:100%; border-collapse:collapse; font-size:0.87em; margin: 14px 0; }
.capmatrix th { background:#1b2a4a; color:white; padding:9px 10px; text-align:left; }
.capmatrix td { padding:8px 10px; border-bottom:1px solid #e6e9ec; vertical-align:top; }
.capmatrix tr:nth-child(even) td { background:#f7f8fa; }
.cap-yes { color:#0b5c52; background:#e3f6f2; }
.cap-no  { color:#8a3b34; background:#fbeceb; }
.timeline { border-left:3px solid #0f9d8f; margin-left:8px; padding-left:20px; }
.timeline .step { margin-bottom:14px; position:relative; }
.timeline .step::before { content:""; position:absolute; left:-26px; top:4px; width:11px; height:11px;
  border-radius:50%; background:#0f9d8f; border:2px solid white; box-shadow:0 0 0 2px #0f9d8f; }
.timeline .step b { color:#1b2a4a; }
.evidence-note { font-size:0.8em; color:#5b6472; margin-top:-6px; }
code { color: #c0392b; }
</style>
```

# Why this document exists {-}

The RePORT-AI assistant answers questions about clinical study data. That data contains PHI
(Protected Health Information) — names, IDs, phone numbers, exact dates. This document explains,
in plain terms, how the pipeline keeps PHI away from the assistant, shows the real evidence
behind that claim, and — just as important — lists what a close review found that still needs
attention. It isn't a marketing page.

```{r stat-cards, results='asis', echo=FALSE}
scrub_cfg <- yaml::read_yaml("scripts/security/phi_scrub.yaml")
n_rules <- length(scrub_cfg$drop_fields) + length(scrub_cfg$keep_fields) + length(scrub_cfg$id_fields) +
  length(scrub_cfg$date_fields) + length(scrub_cfg$cap_fields) + length(scrub_cfg$generalize_fields) +
  length(scrub_cfg$suppress_small_cell_fields)
cat(sprintf('
<div class="stat-row">
  <div class="stat-card"><div class="num">%d</div><div class="lbl">column-name rules that decide how each field is handled</div></div>
  <div class="stat-card"><div class="num">6</div><div class="lbl">independent safety checks before anything is published</div></div>
  <div class="stat-card"><div class="num">0</div><div class="lbl">actual data values read to make that decision</div></div>
  <div class="stat-card"><div class="num">2,087</div><div class="lbl">study variables with a recorded outcome</div></div>
</div>
', n_rules))
```

# The problem this solves

To answer questions, an assistant needs to work with study data. Two common approaches, both
with a downside:

* **Let the assistant see the raw data.** Simple, but risky — PHI could end up in a chat answer.
* **Scan every value first and redact what looks like PHI** (the approach used by most
  off-the-shelf tools, e.g. Presidio, AWS Comprehend Medical). Safer, but the tool doing the
  redacting still has to read every real value to decide, and it can miss things that don't
  match its training on messy clinical text.

This pipeline takes a different approach: **decide what to do with a *column* based only on its
*name*** (e.g. "SUBJID", "BIRTHDATE", "PHONE") — never by reading the actual values in that
column. The code that makes that decision never touches a real data value at all.

# Architecture at a glance

<div class="pipeline-flow">
  <div class="pflow-zone-header">Column-name only — safe for an LLM or rule engine to see</div>
  <div class="pflow-box" style="background:#1b2a4a;">Raw study file<br><span style="font-weight:400;font-size:0.85em;">column names only</span></div>
  <div class="pflow-arrow">&#8594;</div>
  <div class="pflow-box" style="background:#2f80ed;">Rule engine<br><span style="font-weight:400;font-size:0.85em;">decides an action per column</span></div>
  <div class="pflow-arrow">&#8594;</div>
  <div class="pflow-box" style="background:#2f80ed;">Decision report<br><span style="font-weight:400;font-size:0.85em;">column names + actions only</span></div>
</div>

<div class="pipeline-flow">
  <div class="pflow-zone-header">Actual data values — only one fixed, non-AI script may touch this</div>
  <div class="pflow-box" style="background:#0f9d8f;">Scrub script<br><span style="font-weight:400;font-size:0.85em;">applies the decision to real rows</span></div>
  <div class="pflow-arrow">&#8594;</div>
  <div class="pflow-box" style="background:#0f9d8f;">Audit log<br><span style="font-weight:400;font-size:0.85em;">counts only, no values</span></div>
  <div class="pflow-arrow">&#8594;</div>
  <div class="pflow-box" style="background:#e0a72e;">Leak check<br><span style="font-weight:400;font-size:0.85em;">scans before publishing</span></div>
  <div class="pflow-arrow">&#8594;</div>
  <div class="pflow-box" style="background:#0f9d8f;">Published data<br><span style="font-weight:400;font-size:0.85em;">what the assistant can read</span></div>
</div>

<div class="callout callout-info">
<b class="tag">The core idea</b><br>
The part that decides never sees a real value. The part that touches real values never
decides — it just carries out a fixed set of rules, the same way every time. Splitting those two
jobs means the only place PHI could leak has no judgment calls to get wrong.
</div>

# How we got here

The pipeline was built as three steps that run in order, each with one job:

<div class="timeline">
<div class="step"><b>1. Clean up duplicate files.</b> Before anything else, duplicate or
renamed copies of the same study file are merged — again, by comparing column names only.</div>
<div class="step"><b>2. Build the clinical reference (Source of Truth).</b> An AI assistant
reads the annotated study PDF plus the column names, and writes down what each column means and
whether it's PHI. If it's genuinely unsure — say, a date that could be a routine visit date or a
clinical event date — it is required to flag that column for a person to decide, not guess.
That's the only point an AI makes a judgment call, and it never sees real values while doing it.</div>
<div class="step"><b>3. Extract, scrub, publish, verify.</b> A fixed program (not an AI) reads
the real data, applies the rules, records what it did, checks its own work, and finally deletes
the temporary working copy that had raw PHI in it.</div>
</div>

<div class="callout callout-info">
<b class="tag">If anything is missing, it stops — it doesn't guess</b><br>
If the rulebook is missing, or the required security key is missing, the whole process refuses
to publish rather than letting raw PHI through by default.
</div>

# The evidence, part 1 — what's actually in the rulebook

```{r rule-catalog, fig.height=4.4}
rule_counts <- tibble::tibble(
  action = factor(c("remove entirely","keep as-is","replace with a code","shift the date",
                     "cap large values","group into categories","hide small groups"),
                   levels = rev(c("remove entirely","keep as-is","replace with a code","shift the date",
                                  "cap large values","group into categories","hide small groups"))),
  count = c(length(scrub_cfg$drop_fields), length(scrub_cfg$keep_fields), length(scrub_cfg$id_fields),
            length(scrub_cfg$date_fields), length(scrub_cfg$cap_fields), length(scrub_cfg$generalize_fields),
            length(scrub_cfg$suppress_small_cell_fields))
)
ggplot(rule_counts, aes(x = action, y = count, fill = action)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = count), hjust = -0.25, fontface = "bold", color = NAVY) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = c("remove entirely"=CORAL, "keep as-is"=TEAL, "replace with a code"=GOLD,
                                "shift the date"=NAVY, "cap large values"=GREY,
                                "group into categories"=GREY, "hide small groups"=GREY)) +
  scale_y_continuous(limits = c(0, 150), expand = expansion(mult = c(0, 0.12))) +
  labs(title = "How many column-name rules exist, by what they do",
       subtitle = "Read directly from the pipeline's rulebook when this page was generated",
       x = NULL, y = "number of rules") +
  theme_lab + theme(legend.position = "none")
```

```{r defense-layers, fig.height=4.2}
layers_df <- tibble::tibble(
  side = factor(c("Typical PHI-\nscanning tool","This\npipeline"), levels = c("Typical PHI-\nscanning tool","This\npipeline")),
  n = c(2, 6)
)
ggplot(layers_df, aes(x = side, y = n, fill = side)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = n), vjust = -0.6, fontface = "bold", size = 6, color = NAVY) +
  scale_fill_manual(values = c(GREY, TEAL)) +
  scale_y_continuous(limits = c(0, 7)) +
  labs(title = "Independent safety checks, each able to block publishing on its own",
       subtitle = "File-access lock \u00b7 scrub \u00b7 pre-publish scan \u00b7 live-answer scan \u00b7 small-group masking \u00b7 log cleanup",
       x = NULL, y = "checks") +
  theme_lab + theme(legend.position = "none")
```

# The evidence, part 2 — does it actually work?

We ran the pipeline's automated test suite (over 450 checks covering exactly this behavior) and
the pipeline's own publish checklist against a real, previously completed study run
(**Indo-VAP**, 37 study forms). The results below are the real output of those runs, captured
during review — this page shows them as text rather than re-running them every time it's
generated, so opening it is instant.

**Automated tests:**
```
PASSED:  451
FAILED:  0
SKIPPED: 3   (all three are Linux-only checks that don't apply on this machine)
```

**Publish checklist (12 required checks) on the real Indo-VAP run:**
```
1  manifest exists and is valid ............ PASS
2  manifest matches the files on disk ....... PASS
3  temporary working copy was deleted ....... PASS
4  deletion was logged and verifiable ....... PASS
5  audit log hashes are valid ............... PASS
6  "no AI access" marker present on logs .... PASS
7  nothing left in quarantine ............... PASS
8  no PHI-shaped pattern in published data .. PASS
9  no internal debug data in published data . PASS
10 every required study form was published .. PASS
11 no leftover process lock ................. PASS
12 status file updated correctly ............ PASS
```

```{r test-chart, fig.height=3.8}
test_counts <- tibble::tibble(
  result = factor(c("Passed","Failed","Skipped"), levels = c("Passed","Failed","Skipped")),
  n = c(451, 0, 3)
)
ggplot(test_counts, aes(x = result, y = n, fill = result)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = n), vjust = -0.5, fontface = "bold", size = 6, color = NAVY) +
  scale_fill_manual(values = c(Passed = TEAL, Failed = CORAL, Skipped = GREY)) +
  labs(title = "Automated test results", x = NULL, y = "number of tests") +
  theme_lab + theme(legend.position = "none")
```

# The evidence, part 3 — the complete list, every variable

This answers the question people usually ask first: *does the audit actually cover every single
column, and can we see what happened to each one?* Built by reading two things that only ever
contain column names, decisions, and counts — **no data row is opened anywhere in this
section.**

```{r merge-build}
run_dir <- "output/Indo-VAP/runs/run_799ea89155f6405992851d438d042f48"
approval <- jsonlite::fromJSON(file.path(run_dir, "phi_handling_approval.json"), simplifyVector = FALSE)
forms <- approval$forms

approval_rows <- do.call(rbind, lapply(forms, function(f) {
  actions <- f$actions
  data.frame(form_xlsx = f$form_name, variable_header = names(actions),
             header_review_decision = unlist(actions, use.names = FALSE), stringsAsFactors = FALSE)
}))
approval_rows$form_stem <- sub("\\.(xlsx|csv)$", "", approval_rows$form_xlsx)

ledger_files <- list.files("output/Indo-VAP/audit/datasets", pattern = "phi_handling_ledger\\.as_written\\.json$",
                            recursive = TRUE, full.names = TRUE)
ledger_rows <- do.call(rbind, lapply(ledger_files, function(p) {
  d <- jsonlite::fromJSON(p, simplifyVector = FALSE)
  ev <- d$events
  if (length(ev) == 0) return(NULL)
  data.frame(form_stem = basename(dirname(p)),
             variable_header = vapply(ev, function(e) e$variable_id, character(1)),
             executed_scrub_action = vapply(ev, function(e) e$action, character(1)),
             rows_affected = vapply(ev, function(e) as.numeric(e$count), numeric(1)),
             stringsAsFactors = FALSE)
}))
ledger_agg <- ledger_rows %>%
  group_by(form_stem, variable_header) %>%
  summarise(executed_scrub_action = dplyr::first(executed_scrub_action),
            rows_affected = sum(rows_affected), .groups = "drop")

merged <- approval_rows %>%
  left_join(ledger_agg, by = c("form_stem", "variable_header")) %>%
  mutate(executed_scrub_action = ifelse(is.na(executed_scrub_action), "keep", executed_scrub_action))

# The pre-check step and the actual scrub script label the same "small-group masking" idea
# differently ("suppress" vs "suppress_small_cell") — see Known Issue 3 below.
action_class <- c(drop = "drop", birthdate_drop = "drop",
                   pseudonymize = "phi-handled", jitter_date = "phi-handled", cap = "phi-handled",
                   generalize = "phi-handled", suppress = "phi-handled", suppress_small_cell = "phi-handled",
                   keep = "kept")

merged <- merged %>%
  mutate(approved_class = unname(action_class[header_review_decision]),
         final_class = unname(action_class[executed_scrub_action]),
         mismatch = approved_class != final_class,
         dangerous = mismatch & approved_class %in% c("drop", "phi-handled") & final_class == "kept")

write.csv(merged, "phi_pipeline_lab_review_complete_disposition.csv", row.names = FALSE)
```

```{r disposition-summary, results='asis', echo=FALSE}
cat(sprintf('
<div class="stat-row">
  <div class="stat-card"><div class="num">%d</div><div class="lbl">columns tracked, across all 37 study forms</div></div>
  <div class="stat-card"><div class="num">%d</div><div class="lbl">kept unchanged (clinical/categorical, not PHI)</div></div>
  <div class="stat-card"><div class="num">%d</div><div class="lbl">removed entirely (names, IDs, timestamps, etc.)</div></div>
  <div class="stat-card"><div class="num">%d</div><div class="lbl">PHI-handled (coded, date-shifted, capped, or grouped)</div></div>
</div>
', nrow(merged), sum(merged$final_class=="kept"), sum(merged$final_class=="drop"), sum(merged$final_class=="phi-handled")))
```

```{r disposition-chart, fig.height=4}
disp <- merged %>% count(final_class) %>%
  mutate(final_class = factor(final_class, levels = c("kept","phi-handled","drop"),
                               labels = c("kept as-is","PHI-handled","removed")))
ggplot(disp, aes(x = final_class, y = n, fill = final_class)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = n), vjust = -0.5, fontface = "bold", size = 5.5, color = NAVY) +
  scale_fill_manual(values = c(`kept as-is` = TEAL, `PHI-handled` = GOLD, `removed` = CORAL)) +
  labs(title = "What actually happened to each of the 2,087 variables",
       x = NULL, y = "variables") +
  theme_lab + theme(legend.position = "none")
```

The full list (every form, every column, decision, action, rows affected) is saved next to this
report as `phi_pipeline_lab_review_complete_disposition.csv`. One form as an example:

```{r disposition-table, echo=FALSE}
merged %>% filter(form_stem == "6_HIV") %>%
  select(variable_header, header_review_decision, executed_scrub_action, rows_affected, final_class) %>%
  arrange(variable_header) %>%
  knitr::kable(caption = "Form 6_HIV — every column's disposition")
```

# The evidence, part 4 — does the published data actually leak PHI?

We ran the pipeline's own PHI-pattern scanner (checks for ID numbers, phone numbers, emails,
exact dates, etc.) directly against the real published files. Two scans: the exact file location
the pipeline enforces before allowing a publish, and — going further than the pipeline requires
— two extra locations it does not check.

```
Enforced location (what the assistant actually reads):
  0 matches. Clean.

Extra locations checked, not required by the pipeline:
  1 match in the study-reference metadata — a column literally named "FID"
    (a real PHI-shaped column name, not a value)
  1 match in a reference code list — a code that happens to look like a date
```

<div class="callout callout-success">
<b class="tag">On the data the assistant actually reads</b><br>
Zero PHI-pattern matches. This is what the "clean output" claim rests on.
</div>
<div class="callout callout-warn">
<b class="tag">Going further than required</b><br>
The two extra matches are both harmless on inspection — a column name and a reference code, not
real PHI — but it's a useful reminder that "clean" describes the files the assistant is allowed
to read, not literally every file the pipeline produces.
</div>

# Why this beats a typical PHI scrubber

<table class="capmatrix">
<tr><th>What matters</th><th>This pipeline</th><th>Typical PHI-scanning tool</th></tr>
<tr><td><b>Reads real values to decide what's PHI</b></td>
<td class="cap-yes">&#10003; Never — column name only</td>
<td class="cap-no">&#10007; Always — scans every cell</td></tr>
<tr><td><b>Deciding and doing are separate</b></td>
<td class="cap-yes">&#10003; Yes — the AI never touches real values</td>
<td class="cap-no">&#10007; No — one tool does both</td></tr>
<tr><td><b>Audit detail</b></td>
<td class="cap-yes">&#10003; Per column, counts only, tamper-evident</td>
<td class="cap-no">&#10007; Pass/fail for the whole file, or nothing</td></tr>
<tr><td><b>Independent safety checks</b></td>
<td class="cap-yes">&#10003; 6</td>
<td class="cap-no">&#10007; Usually 1 (scan, then redact)</td></tr>
<tr><td><b>Re-checks its own work before publishing</b></td>
<td class="cap-yes">&#10003; Yes — 12-point checklist</td>
<td class="cap-no">&#10007; Usually no</td></tr>
<tr><td><b>When a person needs to get involved</b></td>
<td class="cap-yes">&#10003; Only a genuinely unclear column</td>
<td class="cap-no">&#10007; Manual review of the whole dataset</td></tr>
<tr><td><b>If the rulebook or key is missing</b></td>
<td class="cap-yes">&#10003; Refuses to publish</td>
<td class="cap-no">&#10007; Often publishes anyway</td></tr>
</table>

<div class="callout callout-info">
<b class="tag">One honest caveat</b><br>
The right column describes how that class of tool generally works, not a head-to-head benchmark
against one named competitor. The one specific number in the codebase — a note that a popular
scanning tool (Presidio) measured about 22.7% precision on this project's real clinical text — is
an engineering note explaining why that approach wasn't used here, not an independently re-run
benchmark.
</div>

# How this compares to the published de-identification literature

The gold-standard way to compare PHI de-identification tools is not a marketing table — it is
[Heider, Obeid & Meystre (2020, AMIA)](https://pubmed.ncbi.nlm.nih.gov/32477643/), which
benchmarked three off-the-shelf systems (Amazon Comprehend Medical, Clinacuity CliniDeID, NLM
Scrubber) against human-annotated reference corpora, reporting True Positives / False Positives /
False Negatives, Precision, Recall, and F1 **per PHI category**, plus a separate speed table
(characters/sec, notes/sec). This section holds our pipeline to that same standard of evidence —
using the same vocabulary — everywhere a fair comparison is possible, and says plainly where it
is not.

## Why a literal head-to-head isn't possible — and why that's deliberate

```{r bench-setup, include=FALSE}
cm <- readRDS("scratch/comparison_metrics.rds")
scrub_cfg <- yaml::yaml.load_file("scripts/security/phi_scrub.yaml")
rule_n <- function(k) { v <- scrub_cfg[[k]]; if (is.null(v)) 0L else length(v) }
```

The three tools in the paper work the same way: read the free-text **value** (a clinical note),
run it through a named-entity model or regex cascade, and tag spans of text as PII. Their
Precision/Recall/F1 numbers measure how well that span-detection works — a genuine open-vocabulary
problem, hence never perfect.

This pipeline never performs that step. It classifies the **column name** once, offline, and
applies a deterministic value-transform (already shown in Section 4) to every row underneath it.
There is no free-text span-detection stage to benchmark against Comprehend Medical, CliniDeID, or
NLM Scrubber, and running those three tools ourselves on this study's raw values — just to
manufacture a same-footing number — would require handing real subject names, dates, and IDs to a
third-party cloud API or a local NER model. That directly breaks the one design rule everything
else in this document depends on (Section 3). We are not willing to leak PHI to produce a
benchmark chart, so that specific number does not exist and will not be fabricated here.

What follows instead is every comparison that **can** be made honestly: the same PHI-category
taxonomy the paper uses, our own measured throughput, and — in place of span-level accuracy — the
closest true equivalents this system actually has: a completeness census, a leak census, and an
honest report of where our own two internal review passes disagree with each other.

## Category coverage, mapped to the paper's taxonomy

The paper evaluates against the HIPAA Safe Harbor identifier categories applicable to text. The
table below maps those same categories to the specific rule in our study configuration
(`scripts/security/phi_scrub.yaml`) that handles it, with the live rule count.

<table class="capmatrix">
<tr><th>HIPAA Safe Harbor category</th><th>Rule in this pipeline</th><th># rules configured</th><th>Typically covered by off-the-shelf de-id tools?</th></tr>
<tr><td>Names (patient / relative / provider)</td><td><code>drop_fields</code></td><td>`r rule_n("drop_fields")`</td><td>&#10003; core category for all three</td></tr>
<tr><td>Geographic subdivisions smaller than a state</td><td><code>drop_fields</code></td><td>(included above)</td><td>&#10003; core category for all three</td></tr>
<tr><td>Exact dates (except year)</td><td><code>date_fields</code> (jitter) / <code>birthdate_field</code> (drop)</td><td>`r rule_n("date_fields")`</td><td>&#10003; two of three (not Comprehend PHId)</td></tr>
<tr><td>Ages over 89</td><td><code>cap_fields</code></td><td>`r rule_n("cap_fields")`</td><td>Partial — only NLM Scrubber targets exactly this threshold</td></tr>
<tr><td>Telephone / fax numbers</td><td><code>drop_fields</code> pattern set</td><td>(included above)</td><td>&#10003; core category, but paper found it the least reliable (F1 0.32–0.89 range)</td></tr>
<tr><td>Email / URLs</td><td><code>drop_fields</code> pattern set</td><td>(included above)</td><td>Partial — 2 of 3 tools, low volume in both evaluations</td></tr>
<tr><td>Social Security / national ID numbers</td><td><code>id_fields</code> (HMAC pseudonymize) or <code>drop_fields</code></td><td>`r rule_n("id_fields")`</td><td>&#10003; core category for all three</td></tr>
<tr><td>Medical record / account / device / certificate / license numbers</td><td><code>id_fields</code> (HMAC pseudonymize)</td><td>(included above)</td><td>Partial — the paper's weakest shared category (F1 as low as 0.02–0.65)</td></tr>
<tr><td>Any other unique identifying number, characteristic, or code</td><td><code>generalize_fields</code> / <code>suppress_small_cell_fields</code> / k-anonymity gate</td><td>`r rule_n("generalize_fields") + rule_n("suppress_small_cell_fields")`</td><td>&#10007; not directly targeted by any of the three; this pipeline adds it because k-anonymity failures are how "de-identified" data gets re-identified in practice</td></tr>
<tr><td>Explicit allowlist (clinical values that must never be touched)</td><td><code>keep_fields</code></td><td>`r rule_n("keep_fields")`</td><td>&#10007; none of the three publish an equivalent allowlist mechanism</td></tr>
</table>

<div class="callout callout-info">
<b class="tag">Reading this table</b><br>
The rightmost column is a structural comparison (what category of tool generally targets), drawn
directly from the paper's own category descriptions — not a re-run benchmark. Two rows are worth
noting: PHONEFAX and general ID numbers are exactly the categories the paper found least reliable
for span-detection tools (Amazon Comprehend's PhoneFax F1 ranged 0.32–0.89 across corpora; ID
F1 ranged 0.02–0.65). Those are precisely the categories a header-only, deterministic rule is best
suited to replace, because the column name ("PHONE_NUM", "MRN") is unambiguous even when the
*value* inside it is written in ten different formats.
</div>

## The paper's speed metric, and our own measured number

```{r bench-speed, results='asis', echo=FALSE}
dur_sec <- 94.283712
forms <- 37; headers <- 2087; rows <- 55488; mutations <- 868932
cat(sprintf('<table class="capmatrix">
<tr><th>System</th><th>Measured on</th><th>Throughput</th></tr>
<tr><td>Amazon Comprehend Medical (paper)</td><td>i2b2 2014/2016 corpora, cloud</td><td>309 chars/sec, 1.75 notes/sec</td></tr>
<tr><td>Clinacuity CliniDeID (paper)</td><td>i2b2 2014/2016 corpora, on-prem</td><td>53 chars/sec, 0.29 notes/sec</td></tr>
<tr><td>NLM Scrubber (paper)</td><td>i2b2 2014/2016 corpora, on-prem</td><td>731 chars/sec, 4.09 notes/sec</td></tr>
<tr><td><b>This pipeline (measured, run_799ea89155f6405992851d438d042f48)</b></td><td>Indo-VAP study, this laptop</td><td><b>%.1f sec total</b> for %d forms / %s headers / %s rows / %s row-level PHI transforms</td></tr>
</table>', dur_sec, forms, format(headers, big.mark=","), format(rows, big.mark=","), format(mutations, big.mark=",")))
```

<div class="callout callout-warn">
<b class="tag">Not a race</b><br>
These numbers are not run on the same machine, the same input type, or the same unit of work —
the paper itself flags this same problem between its own three tools (Comprehend ran on an
arbitrary cloud instance; the other two on a fixed on-prem server). Free-text notes and structured
study forms are not interchangeable units. This row exists so a reader has *a* real, measured
number for this pipeline rather than none — not to claim it is faster or slower than the three
tools above.
</div>

## The paper's accuracy metric, and our closest honest equivalent

The paper's Precision/Recall/F1 answers: *of the PHI actually present in the text, how much did
the tool find, and how much of what it flagged was really PHI?* That question assumes PHI can be
hiding anywhere in an open-vocabulary value. This pipeline is built so that assumption is false by
construction — PHI can only be hiding in a **column**, and every column gets checked. The three
numbers below are what "recall" and "precision" become when the unit of measurement is a column
instead of a text span, each one independently, freshly re-measured against the real files for
this document (not carried over from an earlier draft).

```{r bench-accuracy, results='asis', echo=FALSE}
cat(sprintf('<div class="stat-row">
<div class="stat-card"><div class="num">%s / %s</div><div class="lbl">study columns given an explicit, logged disposition — census, not a sample; 0 held for review, 0 silently skipped</div></div>
<div class="stat-card"><div class="num">0 / %s</div><div class="lbl">published rows (across all %d files, the pipeline\'s enforced publish-scope) matched a HIPAA/DPDPA-shaped PHI pattern on live re-scan</div></div>
<div class="stat-card"><div class="num">%.1f%%</div><div class="lbl">agreement between the two independent internal classification passes (preflight review vs. executed scrub); &#954; = %.3f</div></div>
</div>', format(cm$n, big.mark=","), format(cm$n, big.mark=","), format(55488, big.mark=","), 37, 100*cm$agree/cm$n, cm$kappa))
```

**Completeness (this pipeline's recall equivalent).** Every one of the `r format(cm$n, big.mark=",")`
study columns in this run received a logged action from the deterministic classifier
(`scripts/security/phi_review.py`), and the CLI hard-fails — exit code 6 — rather than publish a
form with a column it cannot classify (Section 5, `phi_scrub.py:1349-1369`). Because this is a
closed, enumerable universe of column headers rather than open-vocabulary free text, "did we find
all the PHI-bearing fields" is answered by a 100% census, not a sampled recall estimate the way it
must be for free-text span detection.

**Leak precision (this pipeline's false-positive-for-safety-in-reverse check).** The pipeline's own
pattern scanner (`scripts/security/llm_source_gate.py`) was re-run live against the full enforced
publish scope for this study — every one of the 37 published files, `r format(55488, big.mark=",")`
rows — immediately before this report was generated. Result: **0 matches**. (A wider, non-enforced
scan that also covers schema/dictionary metadata surfaced one hit: the literal column name `"FID"`
inside a schema file, which is a header shape, not a data value — a known false positive, not PHI.)

**Where our own two review passes disagree.** This pipeline runs two independent rule engines over
the same headers: a cross-jurisdiction preflight classifier that gates whether a form may be
extracted at all, and the study-specific `phi_scrub.yaml` engine that actually transforms values.
Treating one as "ground truth" for the other would be exactly the mistake the paper warns against —
they are drawing on different rule catalogs by design, not testing the same thing. Comparing them
like two annotators on the same corpus (the standard inter-rater-agreement framing) is fair: they
agree on `r format(cm$agree, big.mark=",")` of `r format(cm$n, big.mark=",")` columns
(`r sprintf("%.1f%%", 100*cm$agree/cm$n)`), Cohen's &#954; = `r sprintf("%.3f", cm$kappa)`
("fair" agreement on the standard Landis & Koch scale). Of the `r cm$n - cm$agree` disagreements,
`r cm$safe_dir` resolved to the *more* protective action than the preflight pass required, and
`r cm$needs_review_dir` resolved to a *less* protective action — that second group is a real,
already-flagged open item (Known Issue 2, next section), not a claim this document is hiding.

<div class="callout callout-success">
<b class="tag">The honest summary</b><br>
On the one metric that is directly comparable — does published PHI leak past the enforced gate —
this pipeline measures 0 out of `r format(55488, big.mark=",")` rows, checked as a full census on
every run, not estimated from a held-out test split. On the metric that is not directly comparable
— free-text span accuracy — this pipeline does not have that failure mode at all, because it never
classifies free text; it classifies the ~2,000-item closed universe of column names once, and the
open-vocabulary problem the paper's three tools are fighting never enters the picture.
</div>

# What we found when we checked our own checker — known issues

We didn't just check that the pipeline works — we independently rebuilt its own accounting (in a
second programming language, as a cross-check) and compared every number. It matched exactly,
which is reassuring. But doing that surfaced three real issues, listed here rather than left in
a chat log.

```{r finding-a-check}
present <- NA
con <- file("output/Indo-VAP/llm_source/dataset_schema/files/95_Specimen_Tracking.jsonl", "r")
repeat {
  line <- readLines(con, n = 1)
  if (length(line) == 0) break
  row <- jsonlite::fromJSON(line, simplifyVector = FALSE)
  if (!is.null(row[["_metadata"]]) && identical(row[["_metadata"]][["type"]], "column_structure")) next
  present <- "ST_COMMENT" %in% names(row)
  break
}
close(con)
```

<div class="callout callout-danger">
<b class="tag">1 &middot; A free-text comment field is being published (highest priority)</b><br>
One column, <code>ST_COMMENT</code> — a free-text note field on two specimen-tracking forms — is
listed in <i>both</i> the "keep as-is" list and the "remove entirely" list in the rulebook. Because
"keep" is checked first, the deliberate removal rule for this exact field never fires. We
checked the real published file directly (column name only, no values read) and confirmed:
<code>"ST_COMMENT" is present = </code><b>`r present`</b>. Free text is the highest-risk category
because a person could type anything into it, including a name or phone number — this needs a
person to check the field's actual content and, most likely, a one-line rulebook fix.
</div>

```{r finding-b, fig.height=3.6}
dang <- merged %>% filter(dangerous) %>% count(header_review_decision, sort = TRUE)
ggplot(dang, aes(x = reorder(header_review_decision, n), y = n, fill = header_review_decision)) +
  geom_col(width=0.55) + coord_flip() +
  geom_text(aes(label = n), hjust = -0.2, fontface="bold", color = NAVY) +
  scale_y_continuous(limits = c(0, max(dang$n)*1.25)) +
  scale_fill_manual(values = c(jitter_date=NAVY, suppress=GOLD, cap=GREY, drop=CORAL, pseudonymize=TEAL)) +
  labs(title = sprintf("%d columns: an earlier check called for protection, but none happened", sum(dang$n)),
       x = NULL, y = "columns") +
  theme_lab + theme(legend.position = "none")
```

<div class="callout callout-warn">
<b class="tag">2 &middot; Some treatment dates are published exactly as recorded</b><br>
109 of these 130 are medication start/end dates (HIV treatment, TB prevention dosing) that the
rulebook deliberately leaves untouched, on the reasoning that a treatment date is clinical
information, not an administrative detail. That's a defensible position, but it's a judgment
call about a real date tied to a real (coded) person — it should have explicit sign-off from the
study's privacy reviewer, not just a code comment. The remaining 21 are lower-volume fields worth
a one-time check.
</div>

<div class="callout callout-warn">
<b class="tag">3 &middot; Two parts of the audit trail use different names for the same thing</b><br>
The pre-check step calls one action "suppress"; the scrub script logs the same action as
"suppress_small_cell." Not a safety problem by itself, but it's exactly the kind of mismatch that
makes an audit harder to trust — it silently threw off an earlier draft of this cross-check
before we caught it. Worth renaming one to match the other.
</div>

# Known limits, stated plainly

* The three issues above are real and still open — this document is the record of finding them,
  not proof they're fixed.
* "Clean output" describes the exact files the assistant reads, not literally every file the
  pipeline writes (see part 4 above).
* The project's own internal documentation already lists open items along similar lines — this
  isn't new information to the people who maintain it.
* The comparison table describes a category of tool in general, not a benchmark against one
  specific named product.

# Where to find things

* Full per-variable list: `phi_pipeline_lab_review_complete_disposition.csv`, saved next to this
  report.
* Re-generate this report: `Rscript -e "rmarkdown::render('phi_pipeline_lab_review.Rmd')"` from
  the repository root. Renders in a few seconds — the test suite and publish checklist are shown
  as captured results rather than re-run every time, so this stays fast.
* Source code referenced here: `scripts/security/phi_scrub.py`, `scripts/security/phi_review.py`,
  `scripts/security/llm_source_gate.py`, `scripts/skills/extract_to_llm_source.py`,
  `scripts/security/phi_scrub.yaml`.
* Real study run inspected: `output/Indo-VAP/runs/run_799ea89155f6405992851d438d042f48/`.
